---
title: "MetalAnalysis"
author: "Megan Ogburn"
date: "3/6/2020"
output: html_document
---
#Reading in the csv file
Data labels: All field samples begin with the ID label "S" to indicate "sample", then three number spaces to indicate "sample number", followed by "D" for dirt/soil or "C" for caterpillar or "L" for leaf/milkweed. A first leaf sample would have the ID: SOO1L. Soil/dirt samples (plus the Buffalo River sample) were diluted further and re-run, so these samples have an additional "_x10" at the end. 

"Buffalo River" is a sample of known metal concentration which we can use to see if the machine is giving us accurate results. 

"Dirt 1" and "Dirt 2" are control samples of dirt/soil obtained from greenhouse dirt/soil.

"MW 1" and "MW 2" are control samples of leaf/milkweed obtained from greenhouse leaf/milkweed (the numbers of these samples correspond to the numbers on the "Dirt" samples because they were taken from the same pot).

In total there are 47 samples in the data set: dirt x 10 + caterpillar x 10 + leaf x 10 + dirt_x10 x 10 + Dirt 1 + Dirt 2 + MW 1 + MW 2 + Buffalo River + Buffalo River_x10

```{r setup, include=FALSE}

library(tidyverse)
Raw.data<-read.csv("./Data/Raw_lead_data_for_R.csv", header=TRUE)
Raw.data<-as.data.frame(Raw.data)


#Making vectors by sample type
soil.samples<-c("S001D","S002D","S003D","S004D","S005D","S006D","S007D","S008D","S009D","S010D")
soil.controls<-c("Buffalo River","Dirt1","Dirt2")
soil.controls.diluted<-c("Buffalo River_x10")
soil.samples.diluted<-c("S001D_x10","S002D_x10","S003D_x10","S004D_x10","S005D_x10","S006D_x10","S007D_x10","S008D_x10","S009D_x10","S010D_x10")
cata.samples<-c("S001C","S002C","S003C","S004C","S005C","S006C","S007C","S008C","S009C","S010C")
leaf.samples<-c("S001L","S002L","S003L","S004L","S005L","S006L","S007L","S008L","S009L","S010L")
leaf.controls<-c("MW1","MW2")
method.blanks<-c("Method Blank")

#Making data fram of sample types
sample.type.df<-rbind(data.frame(Sample.ID =soil.samples,Sample.type = "soil"),
      data.frame(Sample.ID =soil.controls,Sample.type = "soil.control"),
      data.frame(Sample.ID =soil.controls.diluted,Sample.type = "soil.control.diluted"),
      data.frame(Sample.ID =soil.samples.diluted,Sample.type = "soil.diluted"),
      data.frame(Sample.ID =cata.samples,Sample.type = "cata"),
      data.frame(Sample.ID =leaf.samples,Sample.type = "leaf"),
      data.frame(Sample.ID =leaf.controls,Sample.type = "leaf.control"),
      data.frame(Sample.ID =method.blanks,Sample.type = "method.blank"))

#Merging sample type data frame with Raw.Data data frame by "Sample.ID" while making sure that each df has the same number of levels. Sample.type becomes a column at the end of the Raw.data df
sample.type.df$Sample.ID<-factor(sample.type.df$Sample.ID,levels=levels(Raw.data$Sample.ID))
Raw.data <- Raw.data %>% left_join(sample.type.df,by="Sample.ID")

```

#Visualizing the data

```{r}
str(Raw.data)
```

#Cleaning the data

Removing columns containing units. "X" column gives isotope units which we will not use. All other "X" columns indicate ppb which will be true across all samples. "In115.LR." and "In115.MR" are removed because they wree run as part of the internal standard but unusable for the experiment.

```{r}
#Removing columns containing units. "X" column gives isotope units which we will not use. All other "X" columns indicate ppb which will be true across all samples. "In115.LR." and "In115.MR" are removed because they wree run as part of the internal standard but unusable for the experiment.
Raw.data$X<-NULL
Raw.data$X.1<-NULL
Raw.data$X.2<-NULL
Raw.data$In115.LR.<-NULL
Raw.data$X.3<-NULL
Raw.data$X.4<-NULL
Raw.data$X.5<-NULL
Raw.data$X.6<-NULL
Raw.data$X.6<-NULL
Raw.data$X.7<-NULL
Raw.data$X.8<-NULL
Raw.data$X.9<-NULL
Raw.data$In115.MR.<-NULL
Raw.data$X.10<-NULL


#Creating columns for calculations
samples<-summary(Raw.data$Sample.ID)
names(samples)

#Assigning Final blue volume (in mL) as 15 mL for dirt/soil samples and 5 mL for all other samples (caterpillar, leaf, blank)
Raw.data$Final.Blue.Volume.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples,soil.samples.diluted,soil.controls,soil.controls.diluted), 15, 5))

#Assigning Extraction from blue (in mL) as 0.2 for dirt/soil samples and 2.5 for all other samples (caterpillar, leaf, blank)
Raw.data$Extraction.from.blue.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples,soil.samples.diluted,soil.controls,soil.controls.diluted), 0.2, 2.5))

#Assigning First Orange Volume (in mL) as 10 for dirt/soil samples and 5 for all other samples (caterpillar, leaf, blank)
Raw.data$First.Orange.Volume.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples,soil.samples.diluted,soil.controls,soil.controls.diluted), 10, 5))

#Assigning Extraction from Orange (in mL) as 0.5 for diluted dirt/soil samples and NA for all other samples (caterpillar, leaf, blank, nondiluted dirt/soil)
Raw.data$Extraction.from.Orange.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples.diluted,soil.controls.diluted), 0.5, NA))

#Assigning Second Orange Volume (in mL) as 5 for diluted dirt/soil samples and NA for all other samples (caterpillar, leaf, blank, nondiluted dirt/soil)
Raw.data$Second.Orange.Volume.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples.diluted,soil.controls.diluted), 5, NA))

str(Raw.data)

```

#Calculating PPM for each element

The use of once diluted versus twice diluted data per element is determined based on the raw input from the machine. When the concentration AVG output by the machine is out of the detectable range for that element, use the twice diluted data (elements: lead, vanadium, chromium, manganese, copper zinc; NOT cadmium or nickel). 


##PPM for Cadmium

```{r}
#PPM of Cd in sample for all samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Cd.in.sample=Raw.data$Cd111.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)

#ignore TWICE diluted dirt/soil samples because they are inaccurate
Raw.data$PPM.Cd.in.sample[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type=="soil.control.diluted"]<-NA
```


##PPM for Lead

```{r}
#PPM of Pb in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Pb.in.sample=Raw.data$Pb208.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Pb in sample for twice diluted dirt/soil samples
PPM.Pb.in.diluted.dirt=(Raw.data$Pb208.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Pb.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Pb.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Pb.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA



##Original way I was doing things:
#Raw.data<-Raw.data%>%mutate(PPM.Pb.in.diluted.dirt=Raw.data$Pb208.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))

#Raw.data<-Raw.data%>%mutate(PPM.Pb.in.diluted.dirt=PPM.Pb.in.diluted.dirt/1000)
```


##PPM for Vanadium

```{r}
#PPM of V in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.V.in.sample=Raw.data$V51.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of V in sample for twice diluted dirt/soil samples
PPM.V.in.diluted.dirt=(Raw.data$V51.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.V.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.V.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.V.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


##PPM for Chromium

```{r}
#PPM of Cr in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Cr.in.sample=Raw.data$Cr52.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Cr in sample for twice diluted dirt/soil samples
PPM.Cr.in.diluted.dirt=(Raw.data$Cr52.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Cr.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Cr.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Cr.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


##PPM for Manganese

```{r}
#PPM of Mn in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Mn.in.sample=Raw.data$Mn55.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Mn in sample for twice diluted dirt/soil samples
PPM.Mn.in.diluted.dirt=(Raw.data$Mn55.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Mn.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Mn.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Mn.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


##PPM for Nickel

```{r}
#PPM of Ni in sample for all samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Ni.in.sample=Raw.data$Ni60.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)

#ignore TWICE diluted dirt/soil samples because they are inaccurate
Raw.data$PPM.Ni.in.sample[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type=="soil.control.diluted"]<-NA
```


##PPM for Copper

```{r}
#PPM of Cu in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Cu.in.sample=Raw.data$Cu63.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Cu in sample for twice diluted dirt/soil samples
PPM.Cu.in.diluted.dirt=(Raw.data$Cu63.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Cu.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Cu.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Cu.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


##PPM for Zinc

```{r}
#PPM of Zn in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Zn.in.sample=Raw.data$Zn66.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Zn in sample for twice diluted dirt/soil samples
PPM.Zn.in.diluted.dirt=(Raw.data$Zn66.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Zn.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Zn.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Zn.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


#Seeing only "Concentration AVG"" from "Isotopes" column

```{r}
Concentration.AVG <- Raw.data %>% filter(Isotope=="Concentration AVG") %>% select(Sample.ID, Sample.type, PPM.Cd.in.sample, PPM.Pb.in.sample, PPM.V.in.sample, PPM.Cr.in.sample, PPM.Mn.in.sample, PPM.Ni.in.sample, PPM.Cu.in.sample, PPM.Zn.in.sample)
  

##Original way I was doing things:
# Concentration.AVG$Mass..in.g.of.digested.sample.<-NULL
# Concentration.AVG$Cd111.LR.<-NULL
# Concentration.AVG$Pb208.LR.<-NULL
# Concentration.AVG$V51.MR.<-NULL
# Concentration.AVG$Cr52.MR.<-NULL
# Concentration.AVG$Mn55.MR.<-NULL
# Concentration.AVG$Ni60.MR.<-NULL
# Concentration.AVG$Cu63.MR.<-NULL
# Concentration.AVG$Zn66.MR.<-NULL
# Concentration.AVG$Final.Blue.Volume.in.mL<-NULL
# Concentration.AVG$Extraction.from.blue.in.mL<-NULL
# Concentration.AVG$First.Orange.Volume.in.mL<-NULL
# Concentration.AVG$Extraction.from.Orange.in.mL<-NULL
# Concentration.AVG$Second.Orange.Volume.in.mL<-NULL
# str(Concentration.AVG)
# 
# Concentration.AVG<-as.data.frame(Concentration.AVG)
```


#Seeing only "Concentration STD"" from "Isotopes" column

```{r}
Concentration.STD <- Raw.data %>% filter(Isotope=="Concentration STD") %>% select(Sample.ID, Sample.type, PPM.Cd.in.sample, PPM.Pb.in.sample, PPM.V.in.sample, PPM.Cr.in.sample, PPM.Mn.in.sample, PPM.Ni.in.sample, PPM.Cu.in.sample, PPM.Zn.in.sample)

##Original way I was doing things:
#Concentration.STD=Raw.data[seq(5, nrow(Raw.data), 6), ]
#Concentration.STD$Isotope<-NULL
#Concentration.STD$Mass..in.g.of.digested.sample.<-NULL
#Concentration.STD$Cd111.LR.<-NULL
#Concentration.STD$Pb208.LR.<-NULL
#Concentration.STD$V51.MR.<-NULL
#Concentration.STD$Cr52.MR.<-NULL
#Concentration.STD$Mn55.MR.<-NULL
#Concentration.STD$Ni60.MR.<-NULL
#Concentration.STD$Cu63.MR.<-NULL
#Concentration.STD$Zn66.MR.<-NULL
#Concentration.STD$Final.Blue.Volume.in.mL<-NULL
#Concentration.STD$Extraction.from.blue.in.mL<-NULL
#Concentration.STD$First.Orange.Volume.in.mL<-NULL
#Concentration.STD$Extraction.from.Orange.in.mL<-NULL
#Concentration.STD$Second.Orange.Volume.in.mL<-NULL
#str(Concentration.STD)
#
#Concentration.STD<-as.data.frame(Concentration.STD)
```


#Understanding the data

In his paper in 1999 Mielke found these urban soil PPM 120 Pb, 130 Zn, 3.2 Cd, 138 Mn, 2.1 Cr, 9.8 Ni, 3.8 V, and 12.7 Cu.

I found soil PPMs of: 333 Pb (higher), 46 Zn (lower), 1.7 Cd (lower), 356 Mn (higher), 13 Cr (higher), 21 Ni (higher), 8 V (higher), 46 Cu (higher).

```{r}
#Average leaf concentration
sum.AVG<-Concentration.AVG %>% group_by(Sample.type) %>% summarize(meanCd=mean(PPM.Cd.in.sample),
          meanPb=mean(PPM.Pb.in.sample),
          meanV=mean(PPM.V.in.sample),
          meanCr=mean(PPM.Cr.in.sample),
          meanMn=mean(PPM.Mn.in.sample),
          meanNi=mean(PPM.Ni.in.sample),
          meanCu=mean(PPM.Cu.in.sample),
          meanZn=mean(PPM.Cu.in.sample))
str(sum.AVG)
```
