---
title: "MetalAnalysis"
author: "Megan Ogburn"
date: "3/6/2020"
output: html_document
---
#Reading in the csv file
Data labels: All field samples begin with the ID label "S" to indicate "sample", then three number spaces to indicate "sample number", followed by "D" for dirt/soil or "C" for caterpillar or "L" for leaf/milkweed. A first leaf sample would have the ID: SOO1L. Soil/dirt samples (plus the Buffalo River sample) were diluted further and re-run, so these samples have an additional "_x10" at the end. 

"Buffalo River" is a sample of known metal concentration which we can use to see if the machine is giving us accurate results. 

"Dirt 1" and "Dirt 2" are control samples of dirt/soil obtained from greenhouse dirt/soil.

"MW 1" and "MW 2" are control samples of leaf/milkweed obtained from greenhouse leaf/milkweed (the numbers of these samples correspond to the numbers on the "Dirt" samples because they were taken from the same pot).

In total there are 47 samples in the data set: dirt x 10 + caterpillar x 10 + leaf x 10 + dirt_x10 x 10 + Dirt 1 + Dirt 2 + MW 1 + MW 2 + Buffalo River + Buffalo River_x10

```{r setup, include=FALSE}

library(tidyverse)
Raw.data<-read.csv("./Data/Raw_lead_data_for_R.csv", header=TRUE)
Raw.data<-as.data.frame(Raw.data)


#Making vectors by sample type
soil.samples<-c("S001D","S002D","S003D","S004D","S005D","S006D","S007D","S008D","S009D","S010D")
soil.controls<-c("Buffalo River","Dirt1","Dirt2")
soil.controls.diluted<-c("Buffalo River_x10")
soil.samples.diluted<-c("S001D_x10","S002D_x10","S003D_x10","S004D_x10","S005D_x10","S006D_x10","S007D_x10","S008D_x10","S009D_x10","S010D_x10")
cata.samples<-c("S001C","S002C","S003C","S004C","S005C","S006C","S007C","S008C","S009C","S010C")
leaf.samples<-c("S001L","S002L","S003L","S004L","S005L","S006L","S007L","S008L","S009L","S010L")
leaf.controls<-c("MW1","MW2")
method.blanks<-c("Method Blank")

#Making data frame of sample types
sample.type.df<-rbind(data.frame(Sample.ID =soil.samples,Sample.type = "soil"),
      data.frame(Sample.ID =soil.controls,Sample.type = "soil.control"),
      data.frame(Sample.ID =soil.controls.diluted,Sample.type = "soil.control.diluted"),
      data.frame(Sample.ID =soil.samples.diluted,Sample.type = "soil.diluted"),
      data.frame(Sample.ID =cata.samples,Sample.type = "cata"),
      data.frame(Sample.ID =leaf.samples,Sample.type = "leaf"),
      data.frame(Sample.ID =leaf.controls,Sample.type = "leaf.control"),
      data.frame(Sample.ID =method.blanks,Sample.type = "method.blank"))

#Merging sample type data frame with Raw.Data data frame by "Sample.ID" while making sure that each df has the same number of levels. Sample.type becomes a column at the end of the Raw.data df
sample.type.df$Sample.ID<-factor(sample.type.df$Sample.ID,levels=levels(Raw.data$Sample.ID))
Raw.data <- Raw.data %>% left_join(sample.type.df,by="Sample.ID")

#Left join new excel file "LeadSiteID.csv" by Sample.ID
LeadSiteID<-read.csv("./Data/LeadSiteID.csv", header=TRUE)
LeadSiteID<-as.data.frame(LeadSiteID)
LeadSiteID$Sample.ID<-factor(LeadSiteID$Sample.ID,levels=levels(Raw.data$Sample.ID))
LeadSiteID$Site.ID<-factor(LeadSiteID$Site.ID)

Raw.data <- Raw.data %>% left_join(LeadSiteID, by="Sample.ID")
```

#Visualizing the data

```{r}
str(Raw.data)
```

#Cleaning the data

Removing columns containing units. "X" column gives isotope units which we will not use. All other "X" columns indicate ppb which will be true across all samples. "In115.LR." and "In115.MR" are removed because they wree run as part of the internal standard but unusable for the experiment.

```{r}
#Removing columns containing units. "X" column gives isotope units which we will not use. All other "X" columns indicate ppb which will be true across all samples. "In115.LR." and "In115.MR" are removed because they wree run as part of the internal standard but unusable for the experiment.
Raw.data$X<-NULL
Raw.data$X.1<-NULL
Raw.data$X.2<-NULL
Raw.data$In115.LR.<-NULL
Raw.data$X.3<-NULL
Raw.data$X.4<-NULL
Raw.data$X.5<-NULL
Raw.data$X.6<-NULL
Raw.data$X.6<-NULL
Raw.data$X.7<-NULL
Raw.data$X.8<-NULL
Raw.data$X.9<-NULL
Raw.data$In115.MR.<-NULL
Raw.data$X.10<-NULL


#Creating columns for calculations
samples<-summary(Raw.data$Sample.ID)
names(samples)

#Assigning Final blue volume (in mL) as 15 mL for dirt/soil samples and 5 mL for all other samples (caterpillar, leaf, blank)
Raw.data$Final.Blue.Volume.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples,soil.samples.diluted,soil.controls,soil.controls.diluted), 15, 5))

#Assigning Extraction from blue (in mL) as 0.2 for dirt/soil samples and 2.5 for all other samples (caterpillar, leaf, blank)
Raw.data$Extraction.from.blue.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples,soil.samples.diluted,soil.controls,soil.controls.diluted), 0.2, 2.5))

#Assigning First Orange Volume (in mL) as 10 for dirt/soil samples and 5 for all other samples (caterpillar, leaf, blank)
Raw.data$First.Orange.Volume.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples,soil.samples.diluted,soil.controls,soil.controls.diluted), 10, 5))

#Assigning Extraction from Orange (in mL) as 0.5 for diluted dirt/soil samples and NA for all other samples (caterpillar, leaf, blank, nondiluted dirt/soil)
Raw.data$Extraction.from.Orange.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples.diluted,soil.controls.diluted), 0.5, NA))

#Assigning Second Orange Volume (in mL) as 5 for diluted dirt/soil samples and NA for all other samples (caterpillar, leaf, blank, nondiluted dirt/soil)
Raw.data$Second.Orange.Volume.in.mL<-(ifelse(Raw.data$Sample.ID %in% c(soil.samples.diluted,soil.controls.diluted), 5, NA))

str(Raw.data)

```


#Calculating PPM for Buffalo River Controls

```{r}
#PPM of Cd in Buffalo River Sediment for both no dilution
Raw.data<-Raw.data%>%mutate(PPM.Cd.in.buffalo=Raw.data$Cd111.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)

Raw.data$PPM.Cd.in.buffalo[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type=="soil.control.diluted" | Raw.data$Sample.type== "soil" | Raw.data$Sample.type== "cata"| Raw.data$Sample.type== "leaf.control" | Raw.data$Sample.type== "leaf"]<-NA

#PPM of Pb in Buffalo River Sediment for diluted
Raw.data<-Raw.data%>%mutate(PPM.Pb.in.buffalo=(Raw.data$Pb208.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000)

Raw.data$PPM.Pb.in.buffalo[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type== "soil" | Raw.data$Sample.type== "cata"| Raw.data$Sample.type== "leaf.control" | Raw.data$Sample.type== "soil.control" | Raw.data$Sample.type== "leaf"]<-NA

#PPM of V in Buffalo River Sediment for no dilution
Raw.data<-Raw.data%>%mutate(PPM.V.in.buffalo=Raw.data$V51.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)

Raw.data$PPM.V.in.buffalo[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type=="soil.control.diluted" | Raw.data$Sample.type== "soil" | Raw.data$Sample.type== "cata"| Raw.data$Sample.type== "leaf.control" | Raw.data$Sample.type== "leaf"]<-NA

#PPM of Cr in Buffalo River Sediment for no dilution
Raw.data<-Raw.data%>%mutate(PPM.Cr.in.buffalo=Raw.data$Cr52.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)

Raw.data$PPM.Cr.in.buffalo[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type=="soil.control.diluted" | Raw.data$Sample.type== "soil" | Raw.data$Sample.type== "cata"| Raw.data$Sample.type== "leaf.control" | Raw.data$Sample.type== "leaf"]<-NA

#PPM of Mn in Buffalo River Sediment for diluted
Raw.data<-Raw.data%>%mutate(PPM.Mn.in.buffalo=(Raw.data$Mn55.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000)

Raw.data$PPM.Mn.in.buffalo[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type== "soil" | Raw.data$Sample.type== "cata"| Raw.data$Sample.type== "leaf.control" | Raw.data$Sample.type== "soil.control" | Raw.data$Sample.type== "leaf"]<-NA

#PPM of Ni in Buffalo River Sediment for no dilution
Raw.data<-Raw.data%>%mutate(PPM.Ni.in.buffalo=Raw.data$Ni60.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)

Raw.data$PPM.Ni.in.buffalo[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type=="soil.control.diluted" | Raw.data$Sample.type== "soil" | Raw.data$Sample.type== "cata"| Raw.data$Sample.type== "leaf.control" | Raw.data$Sample.type== "leaf"]<-NA

#PPM of Cu in Buffalo River Sediment for both dilutions
Raw.data<-Raw.data%>%mutate(PPM.Cu.in.buffalo=(Raw.data$Cu63.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000)

Raw.data$PPM.Cu.in.buffalo[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type== "soil" | Raw.data$Sample.type== "cata"| Raw.data$Sample.type== "leaf.control" | Raw.data$Sample.type== "soil.control" | Raw.data$Sample.type== "leaf"]<-NA

#PPM of Zn in Buffalo River Sediment for both dilutions
Raw.data<-Raw.data%>%mutate(PPM.Zn.in.buffalo=(Raw.data$Zn66.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000)

Raw.data$PPM.Zn.in.buffalo[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type== "soil" | Raw.data$Sample.type== "cata"| Raw.data$Sample.type== "leaf.control" | Raw.data$Sample.type== "soil.control" | Raw.data$Sample.type== "leaf"]<-NA

```


#Calculating PPM for each element

The use of once diluted versus twice diluted data per element is determined based on the raw input from the machine. When the concentration AVG output by the machine is out of the detectable range for that element, use the twice diluted data (elements: lead, vanadium, chromium, manganese, copper zinc; NOT cadmium or nickel). 


##PPM for Cadmium

```{r}
#PPM of Cd in sample for all samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Cd.in.sample=Raw.data$Cd111.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)

#ignore TWICE diluted dirt/soil samples because they are inaccurate
Raw.data$PPM.Cd.in.sample[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type=="soil.control.diluted"]<-NA
```


##PPM for Lead

```{r}
#PPM of Pb in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Pb.in.sample=Raw.data$Pb208.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Pb in sample for twice diluted dirt/soil samples
PPM.Pb.in.diluted.dirt=(Raw.data$Pb208.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Pb.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Pb.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Pb.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA



##Original way I was doing things:
#Raw.data<-Raw.data%>%mutate(PPM.Pb.in.diluted.dirt=Raw.data$Pb208.LR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))

#Raw.data<-Raw.data%>%mutate(PPM.Pb.in.diluted.dirt=PPM.Pb.in.diluted.dirt/1000)
```


##PPM for Vanadium

```{r}
#PPM of V in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.V.in.sample=Raw.data$V51.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of V in sample for twice diluted dirt/soil samples
PPM.V.in.diluted.dirt=(Raw.data$V51.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.V.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.V.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.V.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


##PPM for Chromium

```{r}
#PPM of Cr in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Cr.in.sample=Raw.data$Cr52.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Cr in sample for twice diluted dirt/soil samples
PPM.Cr.in.diluted.dirt=(Raw.data$Cr52.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Cr.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Cr.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Cr.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


##PPM for Manganese

```{r}
#PPM of Mn in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Mn.in.sample=Raw.data$Mn55.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Mn in sample for twice diluted dirt/soil samples
PPM.Mn.in.diluted.dirt=(Raw.data$Mn55.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Mn.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Mn.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Mn.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


##PPM for Nickel

```{r}
#PPM of Ni in sample for all samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Ni.in.sample=Raw.data$Ni60.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)

#ignore TWICE diluted dirt/soil samples because they are inaccurate
Raw.data$PPM.Ni.in.sample[Raw.data$Sample.type=="soil.diluted" | Raw.data$Sample.type=="soil.control.diluted"]<-NA
```


##PPM for Copper

```{r}
#PPM of Cu in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Cu.in.sample=Raw.data$Cu63.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Cu in sample for twice diluted dirt/soil samples
PPM.Cu.in.diluted.dirt=(Raw.data$Cu63.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Cu.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Cu.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Cu.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


##PPM for Zinc

```{r}
#PPM of Zn in sample for caterpillar and leaf samples (note, ignore TWICE diluted dirt/soil samples)
Raw.data<-Raw.data%>%mutate(PPM.Zn.in.sample=Raw.data$Zn66.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)/1000)


#PPM of Zn in sample for twice diluted dirt/soil samples
PPM.Zn.in.diluted.dirt=(Raw.data$Zn66.MR.*(Raw.data$Final.Blue.Volume.in.mL/Raw.data$Mass..in.g.of.digested.sample.)*(Raw.data$First.Orange.Volume.in.mL/Raw.data$Extraction.from.blue.in.mL)*(Raw.data$Second.Orange.Volume.in.mL/Raw.data$Extraction.from.Orange.in.mL))/1000

Raw.data$PPM.Zn.in.sample[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]<-PPM.Zn.in.diluted.dirt[Raw.data$Sample.type == "soil.diluted" | Raw.data$Sample.type ==  "soil.control.diluted"]

Raw.data$PPM.Zn.in.sample[Raw.data$Sample.type =="soil" | Raw.data$Sample.type =="soil.control"]<-NA
```


#Seeing only "Concentration AVG"" from "Isotopes" column

```{r}
Concentration.AVG <- Raw.data %>% filter(Isotope=="Concentration AVG") %>% select(Sample.ID, Site.ID, Private.ID, Sample.type, PPM.Cd.in.sample, PPM.Pb.in.sample, PPM.V.in.sample, PPM.Cr.in.sample, PPM.Mn.in.sample, PPM.Ni.in.sample, PPM.Cu.in.sample, PPM.Zn.in.sample)
  

##Original way I was doing things:
# Concentration.AVG$Mass..in.g.of.digested.sample.<-NULL
# Concentration.AVG$Cd111.LR.<-NULL
# Concentration.AVG$Pb208.LR.<-NULL
# Concentration.AVG$V51.MR.<-NULL
# Concentration.AVG$Cr52.MR.<-NULL
# Concentration.AVG$Mn55.MR.<-NULL
# Concentration.AVG$Ni60.MR.<-NULL
# Concentration.AVG$Cu63.MR.<-NULL
# Concentration.AVG$Zn66.MR.<-NULL
# Concentration.AVG$Final.Blue.Volume.in.mL<-NULL
# Concentration.AVG$Extraction.from.blue.in.mL<-NULL
# Concentration.AVG$First.Orange.Volume.in.mL<-NULL
# Concentration.AVG$Extraction.from.Orange.in.mL<-NULL
# Concentration.AVG$Second.Orange.Volume.in.mL<-NULL
# str(Concentration.AVG)
# 
# Concentration.AVG<-as.data.frame(Concentration.AVG)
```


#Seeing only "Concentration STD"" from "Isotopes" column

```{r}
Concentration.STD <- Raw.data %>% filter(Isotope=="Concentration STD") %>% select(Sample.ID, Site.ID, Private.ID, Sample.type, PPM.Cd.in.sample, PPM.Pb.in.sample, PPM.V.in.sample, PPM.Cr.in.sample, PPM.Mn.in.sample, PPM.Ni.in.sample, PPM.Cu.in.sample, PPM.Zn.in.sample)

##Original way I was doing things:
#Concentration.STD=Raw.data[seq(5, nrow(Raw.data), 6), ]
#Concentration.STD$Isotope<-NULL
#Concentration.STD$Mass..in.g.of.digested.sample.<-NULL
#Concentration.STD$Cd111.LR.<-NULL
#Concentration.STD$Pb208.LR.<-NULL
#Concentration.STD$V51.MR.<-NULL
#Concentration.STD$Cr52.MR.<-NULL
#Concentration.STD$Mn55.MR.<-NULL
#Concentration.STD$Ni60.MR.<-NULL
#Concentration.STD$Cu63.MR.<-NULL
#Concentration.STD$Zn66.MR.<-NULL
#Concentration.STD$Final.Blue.Volume.in.mL<-NULL
#Concentration.STD$Extraction.from.blue.in.mL<-NULL
#Concentration.STD$First.Orange.Volume.in.mL<-NULL
#Concentration.STD$Extraction.from.Orange.in.mL<-NULL
#Concentration.STD$Second.Orange.Volume.in.mL<-NULL
#str(Concentration.STD)
#
#Concentration.STD<-as.data.frame(Concentration.STD)
```


#Understanding the data

In his paper in 1999 Mielke found these urban soil PPM 120 Pb, 130 Zn, 3.2 Cd, 138 Mn, 2.1 Cr, 9.8 Ni, 3.8 V, and 12.7 Cu.

I found soil PPMs of: 333 Pb (higher), 474 Zn (higher), 1.7 Cd (lower), 356 Mn (higher), 13 Cr (higher), 21 Ni (higher), 8 V (higher), 46 Cu (higher).

```{r}
#Average leaf concentration
sum.AVG<-Concentration.AVG %>% group_by(Sample.type) %>% summarize(meanCd=mean(PPM.Cd.in.sample),
                                                                  meanPb=mean(PPM.Pb.in.sample),
                                                                  meanV=mean(PPM.V.in.sample),
                                                                  meanCr=mean(PPM.Cr.in.sample),
                                                                  meanMn=mean(PPM.Mn.in.sample),
                                                                  meanNi=mean(PPM.Ni.in.sample),
                                                                  meanCu=mean(PPM.Cu.in.sample),
                                                                  meanZn=mean(PPM.Zn.in.sample))
        
sum.AVG<-sum.AVG %>% filter(Sample.type != "method.blank")  #get rid of the method blank row
tidy.sum.AVG<-sum.AVG %>% gather(metal,concentration,-Sample.type)  #put this back into tidy form


#transfer factor is leaf/soil for Cd and Ni. create a 1 row data.frame with transfer factor for these 2 metals
tf1<-(sum.AVG %>% filter(Sample.type=="leaf") %>% select(meanCd,meanNi))/(sum.AVG %>% filter(Sample.type=="soil") %>% select(meanCd,meanNi))

#transfer factor is leaf/soil.diluted for other metals. create a 1 row data.frame with transfer factor for  other metals
tf2<-(sum.AVG %>% filter(Sample.type=="leaf") %>% select(meanPb,  meanV, meanCr, meanMn,  meanCu, meanZn))/(sum.AVG %>% filter(Sample.type=="soil.diluted") %>% select(meanPb,  meanV, meanCr, meanMn,  meanCu, meanZn))

#use cbind (column bind) to put tf1 and tf2 back together and add a column called Sample type
#NOTE: this puts metals in a different order than they were in before
tf<-cbind(data.frame(Sample.type="transfer factor"),cbind(tf1,tf2))

#trophic.transfer.factor is cata/leaf for all metals 
#just select the metal contrations by using select(-Sample.type) which selects every colums except Sample.type
ttf1<-(sum.AVG %>% filter(Sample.type=="cata") %>% select(-Sample.type))/(sum.AVG %>% filter(Sample.type=="leaf") %>% select(-Sample.type))
ttf<-cbind(data.frame(Sample.type="trophic transfer factor"),ttf1)

#bioaccumulation factor is cata/soil for Cd and Ni
bf1<-(sum.AVG %>% filter(Sample.type=="cata") %>% select(meanCd,meanNi))/(sum.AVG %>% filter(Sample.type=="soil") %>% select(meanCd,meanNi))
  
#bioaccumulation factor is cata/soil.diluted for other metals
bf2<-(sum.AVG %>% filter(Sample.type=="cata") %>% select(meanPb,  meanV, meanCr, meanMn,  meanCu, meanZn))/(sum.AVG %>% filter(Sample.type=="soil.diluted") %>% select(meanPb,  meanV, meanCr, meanMn,  meanCu, meanZn))

#bind bf1 and bf2 together with a column called Sample type to obtain one row for bioaccumulation factor
bf<-cbind(data.frame(Sample.type="bioaccumulation factor"),cbind(bf1,bf2))
                                                
# now use rbind to put  tf, ttf back in sum.AVG
sum.AVG<-rbind(sum.AVG,ttf,tf,bf)   

rm(tf1,tf2,tf,ttf1,ttf,bf1,bf2,bf)  #clean up a bit by getting rid (rm means"remove") of all the temporary data.frames that i created

str(sum.AVG)
```

##Leaf Concentration Histograms

```{r}
#Histogram of Cd leaf concentrations
hist(Concentration.AVG$PPM.Cd.in.sample[Concentration.AVG$Sample.type == "leaf"], main="Histogram of leaf Cd concentrations", xlab="Cd concentration")

#Histogram of Pb leaf concentrations
hist(Concentration.AVG$PPM.Pb.in.sample[Concentration.AVG$Sample.type == "leaf"], main="Histogram of leaf Pb concentrations", xlab="Pb concentration")

#Histogram of V leaf concentrations
hist(Concentration.AVG$PPM.V.in.sample[Concentration.AVG$Sample.type == "leaf"], main="Histogram of leaf V concentrations", xlab="V concentration")

#Histogram of Cr leaf concentrations
hist(Concentration.AVG$PPM.Cr.in.sample[Concentration.AVG$Sample.type == "leaf"], main="Histogram of leaf Cr concentrations", xlab="Cr concentration")

#Histogram of Mn leaf concentrations
hist(Concentration.AVG$PPM.Mn.in.sample[Concentration.AVG$Sample.type == "leaf"], main="Histogram of leaf Mn concentrations", xlab="Mn concentration")
abline(v=20, col='blue') #Mn concentration at or below 20 PPM is a deficient concentration (Marschner 1995)

#Histogram of Ni leaf concentrations
hist(Concentration.AVG$PPM.Ni.in.sample[Concentration.AVG$Sample.type == "leaf"], main="Histogram of leaf Ni concentrations", xlab="Ni concentration")
abline(v=10, col='red') #Ni concentration at or above 10 PPM is a toxic concentration (Marschner 1995)

#Histogram of Cu leaf concentrations
hist(Concentration.AVG$PPM.Cu.in.sample[Concentration.AVG$Sample.type == "leaf"], main="Histogram of leaf Cu concentrations", xlab="Cu concentration")
abline(v=20, col='red') #Cu concentration at or above 20 PPM is a toxic concentration (Marschner 1995)
abline(v=5, col='blue') #Cu concentration at or below 5 PPM is a deficient concentration (Marschner 1995)

#Histogram of Zn leaf concentrations
hist(Concentration.AVG$PPM.Zn.in.sample[Concentration.AVG$Sample.type == "leaf"], main="Histogram of leaf Zn concentrations", xlab="Zn concentration")
abline(v=100, col='red') #Zn concentration at or above 100 PPM is a toxic concentration (Marschner 1995)
abline(v=20, col='blue') #Zn concentration at or below 20 PPM is a deficient concentration (Marschner 1995)
```


##Dirt Concentration Histograms

```{r}
#Histogram of Cd dirt concentrations
hist(Concentration.AVG$PPM.Cd.in.sample[Concentration.AVG$Sample.type == "soil"], main="Histogram of dirt Cd concentrations", xlab="Cd concentration")

#Histogram of Pb dirt concentrations
hist(Concentration.AVG$PPM.Pb.in.sample[Concentration.AVG$Sample.type == "soil.diluted"], main="Histogram of dirt Pb concentrations", xlab="Pb concentration")

#Histogram of V dirt concentrations
hist(Concentration.AVG$PPM.V.in.sample[Concentration.AVG$Sample.type == "soil.diluted"], main="Histogram of dirt V concentrations", xlab="V concentration")

#Histogram of Cr dirt concentrations
hist(Concentration.AVG$PPM.Cr.in.sample[Concentration.AVG$Sample.type == "soil.diluted"], main="Histogram of dirt Cr concentrations", xlab="Cr concentration")

#Histogram of Mn dirt concentrations
hist(Concentration.AVG$PPM.Mn.in.sample[Concentration.AVG$Sample.type == "soil.diluted"], main="Histogram of dirt Pb concentrations", xlab="Mn concentration")

#Histogram of Ni dirt concentrations
hist(Concentration.AVG$PPM.Ni.in.sample[Concentration.AVG$Sample.type == "soil"], main="Histogram of dirt Ni concentrations", xlab="Ni concentration")

#Histogram of Cu dirt concentrations
hist(Concentration.AVG$PPM.Cu.in.sample[Concentration.AVG$Sample.type == "soil.diluted"], main="Histogram of dirt Cu concentrations", xlab="Cu concentration")

#Histogram of Zn dirt concentrations
hist(Concentration.AVG$PPM.Zn.in.sample[Concentration.AVG$Sample.type == "soil.diluted"], main="Histogram of dirt Zn concentrations", xlab="Zn concentration")
```


##Caterpillar Concentration Histograms

```{r}
#Histogram of Cd caterpillar concentrations
hist(Concentration.AVG$PPM.Cd.in.sample[Concentration.AVG$Sample.type == "cata"], main="Histogram of caterpillar Cd concentrations", xlab="Cd concentration")

#Histogram of Pb caterpillar concentrations
hist(Concentration.AVG$PPM.Pb.in.sample[Concentration.AVG$Sample.type == "cata"], main="Histogram of caterpillar Pb concentrations", xlab="Pb concentration")

#Histogram of V caterpillar concentrations
hist(Concentration.AVG$PPM.V.in.sample[Concentration.AVG$Sample.type == "cata"], main="Histogram of caterpillar V concentrations", xlab="V concentration")

#Histogram of Cr caterpillar concentrations
hist(Concentration.AVG$PPM.Cr.in.sample[Concentration.AVG$Sample.type == "cata"], main="Histogram of caterpillar Cr concentrations", xlab="Cr concentration")

#Histogram of Mn caterpillar concentrations
hist(Concentration.AVG$PPM.Mn.in.sample[Concentration.AVG$Sample.type == "cata"], main="Histogram of caterpillar Pb concentrations", xlab="Mn concentration")

#Histogram of Ni caterpillar concentrations
hist(Concentration.AVG$PPM.Ni.in.sample[Concentration.AVG$Sample.type == "cata"], main="Histogram of caterpillar Ni concentrations", xlab="Ni concentration")

#Histogram of Cu caterpillar concentrations
hist(Concentration.AVG$PPM.Cu.in.sample[Concentration.AVG$Sample.type == "cata"], main="Histogram of caterpillar Cu concentrations", xlab="Cu concentration")

#Histogram of Zn caterpillar concentrations
hist(Concentration.AVG$PPM.Zn.in.sample[Concentration.AVG$Sample.type == "cata"], main="Histogram of caterpillar Zn concentrations", xlab="Zn concentration")
```


#Making graphs

```{r}
library(ggplot2)

tidyConc <- Concentration.AVG %>% gather("metal", "concentration", -Sample.ID, -Site.ID, -Private.ID, -Sample.type)

#Graphs for Cd
ggplot(data=tidyConc %>% filter((Sample.type=="leaf" | Sample.type=="cata") & metal=="PPM.Cd.in.sample")) +
  geom_col(aes(x=Site.ID, y=concentration, group=Sample.type, fill=Sample.type), position="dodge") +
  theme(axis.text.x=element_text(angle=90)) 

#Creating scatterplot to check for normal distribution (relationship between X and Y should be linear if normally distributed)
ggplot(data=tidyConc %>% filter((Sample.type=="leaf" | Sample.type=="cata") & metal=="PPM.Cd.in.sample"))+
  geom_point(aes(x=Site.ID, y=concentration, group=Sample.type))


ggplot(data=tidyConc %>% filter((Sample.type=="leaf" | Sample.type=="cata" | Sample.type=="soil.diluted") & metal=="PPM.Cu.in.sample")) +
  geom_col(aes(x=Site.ID, y=concentration, group=Sample.type, fill=Sample.type), position="dodge") +
  theme(axis.text.x=element_text(angle=90))

#test for correlation in Cu between leaves and caterpillars -> because non-normal distribution, a non-parametric test is required
#this assumes leaf and caterpillar and soil samples are in the same order
Cu.leaf <- tidyConc %>% filter((Sample.type=="leaf") & metal=="PPM.Cu.in.sample")
Cu.cata <- tidyConc %>% filter((Sample.type=="cata") & metal=="PPM.Cu.in.sample")
Cu.soil <- tidyConc %>% filter((Sample.type=="soil.diluted") & metal=="PPM.Cu.in.sample")
as.numeric(Cu.leaf$concentration)
cor.test(Cu.leaf$concentration, Cu.soil$concentration) 

#remerge leaf, cata, soil into a dataframe and redo correlation test would be a better way to do this (or using merge command) (left join by site ID)
df<-left_join(Cu.leaf, Cu.cata, by="Site.ID")
cor.test(df$concentration.x, df$concentration.y)

```

##Graphs for Cd

```{r}
library(ggplot2)

#Barchart for Cd
ggplot(data=tidyConc %>% filter((Sample.type=="leaf" | Sample.type=="cata" | Sample.type=="soil") & metal=="PPM.Cd.in.sample")) +
  geom_col(aes(x=Private.ID, y=concentration, group=Sample.type, fill=Sample.type), position="dodge") +
  theme(axis.text.x=element_text(angle=90)) 

#Creating scatterplot to check for normal distribution (relationship between X and Y should be linear if normally distributed) between soil & leaf
ggplot(data=tidyConc %>% filter((Sample.type=="soil" | Sample.type=="leaf") & metal=="PPM.Cd.in.sample"))+
  geom_point(aes(x=Private.ID, y=concentration, group=Sample.type))
  

#Creating scatterplot to check for normal distribution (relationship between X and Y should be linear if normally distributed) between leaf & caterpilar
ggplot(data=tidyConc %>% filter((Sample.type=="leaf" | Sample.type=="cata") & metal=="PPM.Cd.in.sample"))+
  geom_point(aes(x=Private.ID, y=concentration, group=Sample.type))

#Creating scatterplot to check for normal distribution (relationship between X and Y should be linear if normally distributed) between soil & caterpilar
ggplot(data=tidyConc %>% filter((Sample.type=="soil" | Sample.type=="cata") & metal=="PPM.Cd.in.sample"))+
  geom_point(aes(x=Private.ID, y=concentration, group=Sample.type))
```
```

